---
title: "Code Sample"
author: "James Robert Hunter"
lang: en
output:
  html_document:
    toc: true
    toc_depth: 2
    highlight: arrow
---

```{css, echo=FALSE}
/* Accessibility: bold left-aligned captions, dark table text, visible keyboard focus */
caption, .caption, p.caption { color: #333333; font-weight: bold; text-align: left; }
a:focus, [tabindex]:focus { outline: 3px solid #0072B2; outline-offset: 2px; }
.table th, .table td { color: #1a1a1a; }
```

## Research question

Do levels of educational attainment predict wage outcomes?

This analysis uses the `Mroz` dataset from the `{wooldridge}` package, based on married women's labor supply data from the PSID (Mroz, 1987).

```{r setup, message=FALSE, warning=FALSE}
pkgs <- c("tidyverse", "wooldridge", "broom", "quantreg", "sandwich", "lmtest", "patchwork", "scales", "knitr", "kableExtra")
invisible(lapply(pkgs, library, character.only = TRUE)) ; rm(pkgs)
data("mroz", package = "wooldridge") ; raw <- as_tibble(mroz)

opts_chunk$set(dpi = 200, out.width = "100%")

# Table helper
a11y_table <- function(x, caption, digits = 3, ...) {
  out <- kable(x, format = "html", caption = caption, digits = digits, format.args = list(big.mark = ","), ...) |>
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, position = "left")
  out[1] <- gsub("<th ", '<th scope="col" ', out[1], fixed = TRUE)   # mark column headers for screen readers
  out
}

# Plot defaults
blue <- "#0072B2" ; orange <- "#D55E00"               # Okabe-Ito palette (colorblind-safe)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13),
      panel.grid.minor = element_blank(),
      legend.position = "top"))
```

The data contain `r comma(nrow(raw))` observations of `r ncol(raw)` variables: `r paste0("<code>", names(raw), "</code>", collapse = ", ")`.

## Data checks

```{r Missing values}
raw |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(
    everything(),
    names_to = "variable",
    values_to = "n_missing"
  ) |> filter(n_missing > 0) |>
  a11y_table(
    caption = "Table 1. Variables with missing values",
    col.names = c("Variable", "Missing values"))
```

```{r Descriptive statistics}
raw |>
  select(where(is.numeric)) |>
  pivot_longer(everything(), names_to = "variable") |>
  summarise(
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    .by = variable) |>
  a11y_table(
    caption = "Table 2. Descriptive statistics for numeric variables",
    digits = 2,
    col.names = c("Variable", "N", "Mean", "Std. dev.", "Minimum", "Median", "Maximum"))
```

```{r Labor-force participation rate}
lfp_rate <- raw |> summarise(rate = mean(inlf == 1)) |> pull(rate)
```

**Labor force participation rate:** `r percent(lfp_rate, accuracy = 0.1)`

## Analytic sample

```{r Create variables}
df <- raw |>
  mutate(
    employed = as.integer(inlf == 1),
    college = as.integer(educ >= 16),
    log_wage = if_else(                               # `log_wage`: log wage for those in the labor force
      inlf == 1 & lwage > 0,
      lwage,
      NA_real_
    ),
    exper2 = exper^2,                                 # quadratic experience term
    educ_bin = cut(                                   # education categories
      educ,
      breaks = c(0, 8, 11, 12, 15, 21),
      labels = c("<9", "9–11", "12", "13–15", "16+"),
      right = TRUE))

analytic <- df |> filter(!is.na(log_wage))

analytic |>
  summarise(n = n()) |>
  a11y_table(
    caption = "Table 3. Analytic sample size",
    col.names = "Observations")
```

```{r Outlier detection}
iqr_bounds <- analytic |>
  summarise(
    q1 = quantile(log_wage, 0.25),
    q3 = quantile(log_wage, 0.75)
  ) |>
  mutate(lo = q1 - 3 * (q3 - q1), hi = q3 + 3 * (q3 - q1))

n_outliers <- analytic |>
  filter(
    log_wage < iqr_bounds$lo |
      log_wage > iqr_bounds$hi
  ) |> nrow()

tibble(outliers = n_outliers) |>
  a11y_table(
    caption = "Table 4. Log-wage outliers (beyond 3 × IQR from the quartiles)",
    col.names = "Outliers")
```

`r if (n_outliers == 0) "No outliers detected." else paste(n_outliers, "outliers detected.")`

## Figures

```{r Visualize log-wage distribution by college status, fig.width=7, fig.height=5}
educ_colors <- c("College (16+ yrs)" = orange, "No college" = blue)

p1 <- analytic |>
  mutate(
    education = if_else(
      college == 1, "College (16+ yrs)", "No college")) |>
  ggplot(aes(x = log_wage, fill = education)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 25,
    alpha = 0.35,
    position = "identity",
    color = "white") +
  geom_density(
    aes(color = education, linetype = education),      # line type as well as color
    linewidth = 0.9,
    fill = NA) +
  scale_fill_manual(values = educ_colors) +
  scale_color_manual(values = educ_colors) +
  scale_linetype_manual(
    values = c("College (16+ yrs)" = "dashed", "No college" = "solid")) +
  labs(
    title = "Log wage distribution by education",
    x = "Log wage",
    y = "Density",
    fill = NULL,
    color = NULL,
    linetype = NULL)
```

```{r Visualize experience vs. log wage, fig.width=7, fig.height=5}
p2 <- analytic |>
  ggplot(aes(x = exper, y = log_wage)) +
  geom_point(
    alpha = 0.45,
    color = blue) +
  geom_smooth(
    method = "loess",
    formula = y ~ x,
    color = "black",
    se = TRUE) +
  labs(
    title = "Experience vs. log wage (LOESS smoothing)",
    x = "Years of experience",
    y = "Log wage")
```

```{r Visualize education distribution, fig.width=7, fig.height=5}
mean_educ <- mean(analytic$educ)

p3 <- analytic |>
  ggplot(aes(x = educ)) +
  geom_histogram(
    binwidth = 1,
    fill = blue,
    color = "white") +
  geom_vline(
    xintercept = mean_educ,
    linetype = "dashed") +
  annotate(
    "text",
    x = mean_educ + 0.3,
    y = Inf,
    vjust = 1.5,
    hjust = 0,
    label = sprintf("Mean = %.1f yrs", mean_educ)) +
  scale_x_continuous(breaks = seq(4, 18, 2)) +
  labs(
    title = "Education distribution",
    x = "Years of education",
    y = "Count")
```

```{r Visualize mean log wage by education group, fig.width=7, fig.height=5}
wage_by_educ <- analytic |>
  group_by(educ_bin) |>
  summarise(
    n = n(),
    mean_lwage = mean(log_wage),
    .groups = "drop")

p4 <- wage_by_educ |>
  ggplot(aes(x = educ_bin, y = mean_lwage)) +
  geom_col(
    fill = blue,
    width = 0.6) +
  geom_text(
    aes(label = sprintf("%.2f", mean_lwage)),
    vjust = -0.6) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Mean log wage by education group",
    x = "Years of education",
    y = "Mean log wage")
```

```{r Figure alt text}
# Alt text for screen readers
fig_alt <- paste(
  "Four-panel figure.",
  "Top left: density of log wage for women with and without a college degree; the college distribution sits to the right.",
  "Top right: scatterplot of log wage against years of experience, with a LOESS curve that rises over the first 10 years or so and then flattens.",
  "Bottom left: histogram of years of education, with a large spike at 12 years.",
  "Bottom right: bar chart of mean log wage by education group; the values are given in Table 5.")
```

```{r Combined figure, fig.width=11, fig.height=8, fig.alt=fig_alt, fig.cap="Figure 1. Log wages, experience, and education in the analytic sample. The values behind the bottom-right panel are given in Table 5."}
combined_plot <- (p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Labor market outcomes based on Mroz (1987)",
    subtitle = paste("Analytic sample:", comma(nrow(analytic)), "employed married women with positive wages"),
    caption = "Data: Mroz (1987), Panel Study of Income Dynamics, via the {wooldridge} R package.",
    theme = theme(plot.title = element_text(size = 16)))

combined_plot
```

```{r Mean log wage by education group table}
wage_by_educ |>
  a11y_table(
    caption = "Table 5. Mean log wage by education group",
    col.names = c("Years of education", "Observations", "Mean log wage"))
```

## Regression results

```{r OLS regression model using Mincer earnings equation}
# Specify model
ols <- lm(log_wage ~ educ + exper + exper2 + faminc, data = analytic)

term_labels <- c(
  "(Intercept)" = "Intercept",
  educ = "Education (years)",
  exper = "Experience (years)",
  exper2 = "Experience squared",
  faminc = "Family income ($)")

sig4 <- function(x) formatC(x, digits = 4, format = "g")   # 4 significant digits; the `faminc` estimate is tiny

# Robust coefficient table
coeftest(ols, vcov = vcovHC(ols, type = "HC3")) |>
  tidy() |>
  mutate(
    term = term_labels[term],
    across(c(estimate, std.error), sig4),
    p.value = pvalue(p.value)) |>
  a11y_table(
    caption = "Table 6. OLS estimates of log wage with heteroskedasticity-robust (HC3) standard errors",
    align = "lrrrr",
    col.names = c("Term", "Estimate", "Robust std. error", "t statistic", "p-value"))

# Regression output
tidy_ols <- tidy(ols, conf.int = TRUE)

tidy_ols |>
  mutate(
    term = term_labels[term],
    across(c(estimate, std.error, conf.low, conf.high), sig4),
    p.value = pvalue(p.value)) |>
  a11y_table(
    caption = "Table 7. OLS estimates of log wage with conventional standard errors and 95% confidence intervals",
    align = "lrrrrrr",
    col.names = c("Term", "Estimate", "Std. error", "t statistic", "p-value", "95% CI lower", "95% CI upper"))

# Model fit statistics
glance(ols) |>
  select(nobs, r.squared, adj.r.squared, sigma, statistic, p.value, df, df.residual, AIC, BIC) |>
  mutate(p.value = pvalue(p.value)) |>
  a11y_table(
    caption = "Table 8. Model fit statistics",
    align = "r",
    col.names = c(
      "Observations", "R-squared", "Adjusted R-squared", "Residual std. error", "F statistic",
      "p-value", "Model df", "Residual df", "AIC", "BIC"))
```

OLS results suggest that education is positively associated with wages.

```{r Interpretation}
b <- coef(ols) ; peak_exp <- -b["exper"] / (2 * b["exper2"])

tibble(
  interpretation = c("Return to education (%)", "Peak experience (years)"),
  value = c(round(b["educ"] * 100, 2),  round(peak_exp, 1))) |>
  a11y_table(
    caption = "Table 9. Interpretation of the OLS estimates",
    digits = 2,
    col.names = c("Quantity", "Value"))
```
